Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-12-15 paper/paper.tex
Rewrote Section II. Other minor edits.

Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters. Unfortunately, current implementations take several minutes to a day to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate of this model that requires less than 0.3~s to evaluate for a waveform with a starting frequency of 10~Hz (0.04~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods.

@@

%% ________________________________________________________
\section{Introduction}

@@

These results are obtained using Bayesian analyses that compare the data to parameterized waveform models. To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects. This includes descriptions of the point-particle interactions, spins, and tides. Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models. These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5 and 6PN order. Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors [[Favata, YagiYunes, Wade]]. Alternative models exist with more accurate representations of the point particle terms. However, they do not always simultaneously include the spin and tidal effects as well. Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$ [[Favata]]. In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].

One such alternative to the PN waveforms is the effective one body (EOB) formalism (discussed in section~\ref{sec:eob}). It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model. Recently two versions have become available that include both tidal effects and aligned spin. These models are being implemented within the LIGO Algorithm Library (LAL) [[cite]] module \texttt{lalsimulation} under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}. The \texttt{SEOBNRv4T} version [[cite]] used in this paper, includes the tidally induced $\ell=2$ and 3  multipole moments as well as the induced $\ell=2$ and 3 $f$-mode resonances, but does not yet include the spin induced quadrupole moment that can be important for large spins~\cite{Poisson, HarryHinderer2017}. This model also agrees with NR simulations of BNS systems to a level consistent with the numerical accuracy of the simulations, which is less than 1~rad during the last $\sim 10$ orbits before merger~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. The alternative version, \texttt{TEOBResumS} [[cite]], uses a slightly different point-particle and spin prescription. And, in addition, includes the $\ell = 2, 3$, and 4 tidal effects and the spin induced quadrupole moment, but does not include the effect of dynamical tides from the induced $f$-mode resonances. It does, however, use a method for re-summing the expressions for tidal interactions. It also shows similar consistency with NR simulations [[Is this true? cite]].

In addition to accuracy, useful waveform models must be fast enough to be used in Bayesian parameter estimation algorithms such as Markov chain Monte Carlo (MCMC) and Nested Sampling that currently require $10^7$--$10^8$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. Most time-domain waveform implementations such as the PN approximants and the EOB models involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. For PN waveform models, this can take several seconds to minutes, while for the \texttt{SEOBNRv4T} model discussed here, evaluation time can be more than a day for a starting frequency of 10~Hz. Significant work has been done to optimize existing implementations of EOB for BBH systems~\cite{DevineEtienneMcWilliams2016}. However, it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides.

For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates. The first was the stationary phase approximation PN waveform model known as TaylorF2 [[cite]]. Because of the known systematic errors that result from PN waveforms, two additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides} and \texttt{SEOBNRv4\_ROM\_NRTides}. Both of these models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017}; at low frequencies, the tidal effect is that of TaylorF2. It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins. The point-particle and spin prescriptions were different. For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model [[cite]]. It uses TaylorF2 at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies. The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM} [[cite]] to the EOB waveform \texttt{SEOBNRv4} for BBH systems [[cite]]. Both of these models are improvements to \texttt{TaylorF2}, but do not incorporate all known physical effects in a consistent way. We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of waveform models. The essential idea is to construct an interpolating function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the parameter space of waveforms using a reduced basis of orthonormal functions whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. The two most common approaches have been singular value decomposition (SVD)~\cite{Puerrer2014, Puerrer2015} or a greedy method that selects a subset of the training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. In this work, instead of using a reduced basis to efficiently represent the EOB waveform as a function of frequency, we will instead use the fact that the frequency-domain waveform is approximately known analytically via the TaylorF2 waveform. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we simply choose to represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

This solves the problem of efficiently representing waveforms as a function of frequency, but leaves the question of interpolating as a function of waveform parameters $\bx$ unanswered. Previous work on surrogate models have focused mainly on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation will be unfeasable. Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on an interpolation method known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}. GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016} and to estimate noise power spectral densities [[cite]].

This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi et al.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. It also enables additional techniques such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} and multi-banding~\cite{VinciguerraVeitchMandel2017} to accelerate likelihood evaluation. The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.

We organize the paper as follows. In Section II we provide an overview of the EOB waveform model used for constructing the surrogate and the approximations we use to reduce the dimensionality of the parameter space. In section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in section IV. Then, finally discuss future improvements and applications in Section V.

\textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$. To agree with the GPR literature, we will refer to the waveform parameters as $\bx$ and the hyperparameters in the GPR model as $\btheta$.

@@

The effective-one-body (EOB) approach to the general-relativistic 2-body problem, first described in Ref.~\cite{Buonanno:1998gg}, has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of aligned-spin binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to 0.98 for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter---BNS or neutron-star-black-hole (NSBH) binaries. The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion---be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of the f-mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.

%We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}\blue{[[Which is the Reference?]]}.

More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs.\footnote{This model is implemented in the LIGO Algorithm Library as SEOBNRv4.} The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning. This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.

@@

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below. Let $A$ and $B$ be the body labels. Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.  Explicitly, including quadrupolar and octupolar tidal effects, we have
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
\end{align}
\end{widetext}
where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$ and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star. Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$.\footnote{Here $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.} This enhancement factor is 1 for low orbital frequencies, and increases to about 2 as $\ell$ times the orbital frequency approaches the $f$-mode frequency.

Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
\end{equation} 
where the point-mass piece $h_{\ell m}^{\textrm{pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and the tidal piece $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}. The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the GW flux (Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived.

>>>>

Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as the effect of dynamical tides, and the formalism is accurate enough to avoid significant biases in the measured parameters. Unfortunately, current implementations take several minutes to a day to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate of this model that requires less than 0.3~s to evaluate for a waveform with a starting frequency of 10~Hz (0.04~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with $\sim 1000$ waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods. \red{[[Mismatch is ...]]}

@@

%% ________________________________________________________
\red{Choose point-mass or point-particle.}
\section{Introduction}

@@

These results are obtained using Bayesian analyses that compare the data to parameterized waveform models. To avoid systematic errors highly accurate models are needed, and they must include all relevant physical effects. This includes descriptions of the point-particle interactions, spins, and tides. Prior to the discovery of GW170817, most parameter estimation analyses were done using post-Newtonian (PN) waveform models. These models are currently known up to 3.5 PN order in the point-particle and spin terms, and the leading and next-to-leading order terms for tidal interactions that enter at 5 and 6PN order. Unfortunately, the lack of point-particle terms above 3.5PN order leads to systematic errors in the tidal parameters that are as large as the statistical errors~\cite{Favata2014, YagiYunes2014, WadeCreightonOchsner2014}. Alternative models exist with more accurate representations of the point particle terms. However, they do not always simultaneously include the spin and tidal effects as well. Not including spin effects, for example, can bias the measured tidal parameter for even moderate dimensionless spins of $|\chi_i| \le 0.04$~\cite{Favata2014}. In addition, ignoring tidal interactions can bias the measured mass ratio from equal mass to unequal mass [[cite who?]].

One such alternative to the PN waveforms is the effective one body (EOB) formalism discussed in section~\ref{sec:eob}. It includes several dynamical effects beyond the standard PN adiabatic inspiral evolution and can also be tuned to numerical relativity (NR) simulations. Until recently, however, no version of the EOB formalism included the effects of spin and tides in the same model. Recently two versions have become available that include both tidal effects and aligned spin. These models are being implemented within the LIGO Algorithm Library (\texttt{LAL})~\cite{lal} under the names \texttt{SEOBNRv4T} and \texttt{TEOBResumS}. The \texttt{SEOBNRv4T} model~\cite{Hinderer:2016eia,Steinhoff:2016rfi} used in this paper, includes the tidally induced $\ell=2$ and 3  multipole moments as well as the induced $\ell=2$ and 3 $f$-mode resonances, but does not yet include the spin induced quadrupole moment that can be important for large spins~\cite{Poisson1998, HarryHinderer2017}. This model also agrees with NR simulations of BNS systems to a level consistent with the numerical error of the simulations (less than 1~rad during the last $\sim 10$ orbits before merger)~\cite{DietrichHinderer2017, KiuchiKawaguchiKyutoku2017}. The alternative model, \texttt{TEOBResumS}~\cite{Nagar2017}, uses a slightly different point-particle and spin prescription. For tidal interactions, the model includes the $\ell = 2, 3$, and 4 terms and uses a method for re-summing these expressions. It does not include the effect of dynamical tides from the induced $f$-mode resonances, but does include the spin induced quadrupole moment.

%The alternative version, \texttt{TEOBResumS}~\cite{Nagar2017}, uses a slightly different point-particle and spin prescription. And, in addition, includes the $\ell = 2, 3$, and 4 tidal effects and the spin induced quadrupole moment, but does not include the effect of dynamical tides from the induced $f$-mode resonances. It does, however, use a method for re-summing the expressions for tidal interactions.

In addition to accuracy, useful waveform models must be fast enough to be used in Bayesian parameter estimation algorithms such as Markov chain Monte Carlo (MCMC) and Nested Sampling that currently require $10^7$--$10^8$ sequential waveform evaluations. Waveform evaluation times must therefore be significantly less than 1~s in order to run in less than a month for each event. Most time-domain waveform implementations such as the PN approximants and the EOB models involve solving a set of ordinary differential equations and then Fourier transforming the results into the frequency domain. For PN waveform models, this can take several seconds for a starting frequency of 10~Hz, while for the \texttt{SEOBNRv4T} model discussed here, the equivalent evaluation time can be a day. Significant work has been done to optimize existing implementations of EOB for binary black hole (BBH) systems by a factor of several hundred~\cite{DevineEtienneMcWilliams2016}. However, it is not clear that sufficient speed ups can be achieved for the BNS model discussed here with dynamical tides.

For the GW170817 event, therefore, a set of frequency-domain analytic or semi-analytic waveform models were used as templates. The first was the stationary phase approximation PN waveform model known as \texttt{TaylorF2}~\cite{BuonannoIyerOchsner2009}. Because of the known systematic errors that result from PN waveforms, two additional models were produced for comparison: \texttt{IMRPhenomD\_NRTides} and \texttt{SEOBNRv4\_ROM\_NRTides}. Both of these models used an identical prescription for tides presented in Ref.~\cite{DietrichBernuzziTichy2017}. At low frequencies, the tidal effect is that of \texttt{TaylorF2}. It then transitions to a fit to the tidal effect measured by a small set of numerical simulations with varying mass ratio and aligned-spins. The point-particle and spin prescriptions were different. For \texttt{IMRPhenomD\_NRTides}, the waveform is based on the phenomenological \texttt{IMRPhenomD} BBH model~\cite{HusaKhanHannam2016, KhanHusaHannam2016}. It uses \texttt{TaylorF2} at low frequencies then transitions to fits to the EOB waveform at mid frequencies and fits to NR simulations at high frequencies. The \texttt{SEOBNRv4\_ROM\_NRTides} model is based on the fast surrogate model \texttt{SEOBNRv4\_ROM}~\cite{Puerrer2015} for the EOB waveform \texttt{SEOBNRv4} for BBH systems~\cite{Bohe:2016gbl}. Both of these models are improvements to \texttt{TaylorF2}, but do not incorporate all known physical effects in a consistent way. We therefore build a fast surrogate that reproduces as accurately as possible \texttt{SEOBNRv4T} in the frequency domain.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of waveform models. The essential idea is to construct an interpolating function that approximates a waveform model as a function of frequency (or time) and waveform parameters $\bx$. Previous works have focused on efficiently representing the parameter space of waveforms using a reduced basis of orthonormal functions whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. The two most common approaches have been singular value decomposition~\cite{Puerrer2014, Puerrer2015} and a greedy method that selects a subset of training-set waveforms then orthonormalizes them with the Graham-Schmidt process~\cite{FieldGalleyHesthaven2014, LackeyBernuzziGalley2017, BlackmanFieldGalley2015, BlackmanFieldScheel2017a, BlackmanFieldScheel2017b}. In this work, instead of using a reduced basis to efficiently represent the EOB waveform as a function of frequency, we will instead use the fact that the frequency-domain waveform is approximately known analytically via the TaylorF2 waveform. We then build a surrogate of the difference (or residual) between the EOB waveform and the TaylorF2 waveform. Because this residual is small, extremely high accuracy is not required, and we simply choose to represent the residual as a function of frequency using cubic splines between a small set of frequency nodes.

The final step is to interpolate between waveform parameters $\bx$. Previous work on surrogate models focused on parameter spaces with three or fewer dimensions. This allows one to approximate the waveform as a function of $\bx$ using standard interpolation techniques such as tensor spline~\cite{Puerrer2014, Puerrer2015} or Chebyshev interpolation~\cite{LackeyBernuzziGalley2017}. These interpolation techniques typically require waveforms to be evaluated on a rectangular grid, and thus suffer from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the problem here, we will only be able to reduce the parameter space to five dimensions, so grid based interpolation is unfeasible. Recent work on optimally choosing waveforms for both EOB~\cite{DoctorFarrHolz2017} and NR~\cite{BlackmanFieldScheel2017a, BlackmanFieldScheel2017b} waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

In this paper we focus on a technique known as Gaussian process regression (GPR)~\cite{RasmussenWilliams2006} that does not require a regular grid. Importantly, GPR also provides a convenient estimate of its own uncertainty. This allows us to iteratively add new points to the training set in such a way as to minimize the interpolation error over the parameter space for a given number of waveforms in the training set. Similar approaches have recently been used to optimally sample waveforms for a 2-dimensional aligned spin BBH surrogate using GPR~\cite{DoctorFarrHolz2017} as well as for a surrogate of eccentric BBH mergers~\cite{HuertaMooreKumar2017}. GPR has also been used in GW data analysis to marginalize over waveform uncertainties in parameter estimation~\cite{MooreGair2014, MooreBerryChua2016}.

This work is a continuation of the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015} and multi-banding~\cite{VinciguerraVeitchMandel2017}. The resulting surrogate is the first EOB waveform model that consistently incorporates spin and tides and is fast enough for production parameter estimation codes.

We organize the paper as follows. In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in section IV. Finally, we discuss applications and future improvements in Section V.

\textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$.

@@

The EOB approach to the general-relativistic 2-body problem, first described in Ref.~\cite{Buonanno:1998gg}, has proven successful in modeling the dynamics and GW emission of compact binaries. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of NR simulations of aligned-spin BBH systems for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to 0.98 for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}.

This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu} for binaries that contain NSs. The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion. The forcing tidal field varies at a multiple of the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of the $f$-mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.

More precisely, spin effects are those of the point-mass model \texttt{SEOBNRv4}~\cite{Bohe:2016gbl} that was calibrated to 141 NR simulations of BBH systems with aligned spins. The model, however, does not account for the spin-induced quadrupole moment~\cite{Poisson1998}. This effect appears at 2PN order as compared to the tidal effects that first appear at 5PN order, but is only significant for systems with large spin. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{HarryHinderer2017}.

@@

Dynamical tidal effects are implemented in the EOB model through a modification of the potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. We adopt the tidally-augmented expression for $\Delta_u$ discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}:
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass EOB term (Eq.~(2.2) of Ref.~\cite{Bohe:2016gbl}), and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. If either component of the binary is a black hole, then we set the tidal polarizabilities to zero. Let $M_{A,B}$ be the masses bodies $A$ and $B$, and $M = M_A + M_B$ be the total mass of the binary. The tidal contribution, including quadrupolar and octupolar dynamical tides, is
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,.
\end{align}
\end{widetext}
Here, $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$, and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\hat{\omega}_{0\ell}^{A,B}$\footnote{Here, $\hat{\omega}_{0\ell}^{A,B}=M_{A,B}\omega_{0\ell}^{A,B}$, where $\omega_{0\ell}^{A,B}$ is the frequency in geometrized units.}. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the NS compactness, which depends on the NS radius $R_{A,B}$. Here, $\hat{k}_{\ell\,\textrm{dyn}}(u;\hat{\omega}_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor (Eq.~(11) of Ref.~\cite{Dietrich:2017feu}), which depends on the value of the $f$-mode $2^\ell$-pole dimensionless angular frequency, $\hat{\omega}_{0\ell}^{A,B}$. This enhancement factor is 1 for low orbital frequencies, and increases to $\sim 2$ as $\ell$ times the orbital frequency approaches the $f$-mode frequency.
